### 2022-02-14 chapters/intro/intro.tex
encore plus de refs
relecture I

Nevertheless, FCR has been heavily impacted by the human activities of the major metropolitan area of greater Miami. These impacts include densely populated coastlines, large numbers of visitors, polluted terrestrial run-off and overfishing \citep{jackson2014status}. For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to have caused the death of $>$560,000 corals \citep{cunning2019extensive}. Florida's coral reefs have therefore 
d significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}. Populations of previously dominant reef-building acroporids almost totally disappeared throughout the Caribbean and Florida \citep{aronson2001white}, and coral cover dropped from 40 to 60\% in 1975 to $<$7\% in the Florida Keys \citep{jackson2014status} and $<$3\% in the northern section of the FRT \citep{walton2018impacts}. Furthermore, since 2014, many coral species have experienced widespread and severe decline caused by the ongoing outbreak of stony coral tissue loss disease (SCTLD). For instance, populations of \textit{M. cavernosa}, once one of the most abundant species of the FRT, decreased by 45\% in abundance between 2015 and 2016 \citep{walton2018impacts}.

@@

Additionally, Florida is a prime landfall target for tropical cyclones from June through November (Fig. \ref{inro:landfall}). Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds $>$ 209 km/h) \citep{neumann1999tropical}. Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}. This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}. Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,gardner2005hurricanes,aronson2001white}. They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects, tweel2014contribution}. Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}. Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}. This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{hendee1998champ}. Hurricane might therefore interrupt larval exchanges between otherwise connected reefs but also create new connectivity pathways. As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricanes on wave-induced processes could inform the management of reefs under future climatic conditions. Furthermore, as incoming storm-generated waves dissipate over reefs, better understanding the protective role of coral reefs might promote their restoration as tool to reduce coastal flooding hazard \citep{roelvink2021coral}.

@@

\section{Modeling hydrodynamics and transport processes in the FRT}
The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean. The large-scale circulation in this region is dominated by the interplay of the Florida Current (FC) and the Loop Current (LC). The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC). The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream (Fig. \ref{intro:gom}). The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}. The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT. These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability. The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}. These eddies significantly impact the ecology of the southwest Florida Shelf by causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}. Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}. These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}. However, on the northernmost section of the FRT, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}. The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}

@@

The aforementioned large-scale ocean circulation significantly impact transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation features. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. Unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way. They allow to locally refine the model resolution to $\sim~100$ m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}. In contrast, the best resolution currently available in the Florida Keys using structured-grid models is $\sim900$ m (FKeyS-HYCOM, \citealp{kourafalou2012florida}). Nested grid models are another possible approach \citep{warner2010development} but they lack the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.

In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}). This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}. Moreover, by coupling SLIM with the HYbrid Coordinate Ocean Model (HYCOM-GOMl0.04, \citealp{chassignet2007hycom}), SLIM was able to indirectly represent the baroclinic dynamics of the FC. Using this coupling, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs. In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents. To answer our different research questions, we adapted this larval transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.

>>>>

Nevertheless, FCR has been heavily impacted by the human activities of the major metropolitan area of greater Miami. These impacts include densely populated coastlines, large numbers of visitors, polluted terrestrial run-off and overfishing \citep{jackson2014status}. For example, dredging operations during the PortMiami Deep Dredge Project (PMDDP) was reported to have caused the death of $>$560,000 corals \citep{cunning2019extensive}. Florida's coral reefs have therefore declined significantly over the past decades due to anthropogenic stressors and numerous disease outbreaks \citep{gardner2003long, jackson2014status}. Populations of previously dominant reef-building acroporids almost totally disappeared throughout the Caribbean and Florida \citep{aronson2001white}, and coral cover dropped from 40 to 60\% in 1975 to $<$7\% in the Florida Keys \citep{jackson2014status} and $<$3\% in the northern section of the FRT \citep{walton2018impacts}. Furthermore, since 2014, many coral species have experienced widespread and severe decline caused by the ongoing outbreak of stony coral tissue loss disease (SCTLD). For instance, populations of \textit{M. cavernosa}, once one of the most abundant species of the FRT, decreased by 45\% in abundance between 2015 and 2016 \citep{walton2018impacts}.

@@

Additionally, Florida is a prime landfall target for tropical cyclones from June through November (Fig. \ref{inro:landfall}). Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds $>$ 209 km/h) \citep{neumann1999tropical}. Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}. This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}. Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,gardner2005hurricanes,aronson2001white}. They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects, tweel2014contribution}. Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}. However, these upwelling also reduce sea surface temperature and can thus alleviate the thermal stress experienced by corals \citep{wilkinson2008status}. Finally, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}. This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{hendee1998champ}. Hurricane might therefore interrupt larval exchanges between otherwise connected reefs but also create new connectivity pathways. As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricanes on wave-induced processes could inform the management of reefs under future climatic conditions. Furthermore, as incoming storm-generated waves dissipate over reefs, better understanding the protective role of coral reefs might promote their restoration as tool to reduce coastal flooding hazard \citep{roelvink2021coral}.

@@

\section{Modeling hydrodynamics and transport processes in the FRT}
The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean. The large-scale circulation in this region is dominated by the interplay of the Florida Current (FC) and the Loop Current (LC). The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC). The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream (Fig. \ref{intro:gom}). The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}. The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT. These variations of the FC position have been shown to have no seasonal pattern, and a strong interannual variability. The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}. These eddies significantly impact the ecology of the southwest Florida Shelf by causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution,limouzy1997translocation,sponaugle2005florida, kourafalou2012florida}. Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}. These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}. However, on the northernmost section of the FRT, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}. The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}

@@

The aforementioned large-scale ocean circulation significantly impacts transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation features. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. A reference ocean model of the FRT is the 3D HYbrid Coordinate Ocean Model (HYCOM, \citealp{chassignet2007hycom}), or more specifically the nested high resolution model FKEYS-HYCOM, that covers the Florida Straits, South Florida and Keys with a $\sim$900 m resolution \citep{kourafalou2012florida}. This model is embedded within the coarser GoM-HYCOM, that models the whole GoM and Florida Straits using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}. However as a structured-grid model, the finer resolution of HYCOM is limited by computational resources and the nested grid approach lacks the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}. In contrast, unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way. They allow to locally refine the model resolution to $\sim~100$ m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.

In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}). This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}. Moreover, by coupling SLIM with GOM-HYCOM, the model was able to indirectly represent the baroclinic dynamics of the FC. Using this coupling, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs. In this study, coral larvae were modeled using a Lagrangian particle tracking model \citep{van2018lagrangian} forced by SLIM currents. To answer our different research questions, we adapted this larval transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.

### 2022-02-14 include/abstract.tex
relecture I

Coral populations have declined dramatically worldwide under the combined effects of climate change and local anthropogenic stressors. In addition to ocean warming and acidification, threats to coral reefs in the Caribbean include frequent and virulent disease outbreaks, and intensifying hurricanes. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of the PortMiami Deep Dredge Project (PMDDP), the disease has now spread through the entire FCR and has been reported in numerous territories of the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached its southwestern end, between the Marquesas and the Dry Tortugas (DRTO). Although the causative agent of the disease remains unknown, studies showed that its transmission was likely waterborne and that sediments could act as SCTLD vectors. The hydrodynamics should therefore be highly explanatory of its spread. Furthermore, Florida is a prime landfall target for hurricanes during the reproduction period of corals. In addition to causing wholesale destruction of the reefs, hurricanes might thus also impact larval dispersal through wind-wave-induced currents. Here, we use the high resolution coastal ocean model SLIM to capture transport processes in the FCR, and study the impact of SCTLD and hurricanes to coral reefs. First, a coupled epidemio-hydrodynamic model is developed to reproduce and understand the observed spread of SCTLD in Florida. A sediment transport model is then used to evaluate the impact of the PMDDP on the observed onset of the disease in 2014. Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys. Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current-Florida Current system. Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease. Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes. This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as the FCR.

>>>>

Coral populations have declined dramatically worldwide under the combined effects of climate change and local anthropogenic stressors. In addition to ocean warming and acidification, threats to coral reefs in the Caribbean include frequent and virulent disease outbreaks, and intensifying hurricanes. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of the PortMiami Deep Dredge Project (PMDDP), the disease has now spread through the entire FCR and has been reported in numerous territories of the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached its southwestern end, between the Marquesas and the Dry Tortugas (DRTO). Although the causative agent of the disease remains unknown, studies showed that its transmission was likely waterborne and that sediments could act as SCTLD vectors. The hydrodynamics should therefore be highly explanatory of its spread. Furthermore, Florida is a prime landfall target for hurricanes during the reproduction period of corals. In addition to causing wholesale destruction of the reefs, hurricanes might thus also impact larval dispersal through wind-wave-induced currents. Here, we use the high resolution coastal ocean model SLIM to capture transport processes in the FCR, and study the impact of SCTLD and hurricanes to coral reefs. First, a coupled epidemio-hydrodynamic model is developed to reproduce and understand the observed spread of SCTLD in Florida. A sediment transport model is then used to evaluate the impact of the PMDDP on the observed onset of the disease in 2014. Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys. Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current/Florida Current system. Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease. Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes. This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as the FCR.
